On Sample Size for Block Bootstrap Confidence Intervals to Have Desired Coverage Rates

### 2023-08-22 Manuscript/manuscript.tex
minor grammar fixes and new paper reference

The goal of this paper is to provide recommendations on necessary sample size
for block bootstrap with dependent data, similar to what was done for basic
bootstrap in \citet{hesterberg2015teachers}. We consider a simple situation of
a stationary time series, where the parameters of interests are the mean,
standard deviation, and the first-order autocorrelation coefficient. We
compare six variants of block bootstrap CIs from the literature
\citep{diciccio1996bootstrap, rice2006mathematical}: a standard normal CI, and
a Student's $t$ CI, a percentile CI, a bias-corrected CI, a bias-correct and
accelerated ($BC_a$) CI, and a CI proposed in this article. Their empirical
coverage rates at different sample sizes and dependence levels are compared in
a simulation study. The results of this study suggest that recovery of
temporal dependence parameters is reliant on the type of interval used.

@@

\paragraph{Student's $t$ CI}
The procedure for constructing a Student's $t$ CI based on standard bootstrap
is described in \citet[p.158]{efron1993introduction}. Let $t_{(\alpha, k)}$ be
the upper $\alpha$ quantile of a $t$ distribution with $k$ degrees of freedom.
With block bootstrapping, a $(1 - \alpha)100\%$ Student's $t$ CI is
\[
(\hat{\theta}_{n} - t_{(1-\alpha/2), k - 1}\hat{\text{SE}}, \quad
\hat{\theta}_{n} - t_{(\alpha/2), k -1}\hat{\text{SE}}),
\]
\eds{define $t_{(.)}$ and check the signs}  \mc{addressed}
where $k$ is the number of blocks. This CI is centered by the point estimator
$\hat\theta_n$ and is symmetric. Like the standard normal interval, the
Student's $t$ CI is a table-based bootstrap interval approach. In this case,
its validity relies on whether the distribution of $\hat\theta_n$ is
reasonably well approximation by the $t_{k-1}$ distribution with an expected
value of $\theta$, and whether the bootstrap $\widehat{\text{SE}}$
approximates the true standard error.

@@

\paragraph{Bias Corrected and Accelerated ($BC_a$) Bootstrap CI}
The $BC_a$ CI was first suggested in \citet{efron1987better}.Let
$\hat{z}_0 = \Phi^{-1}\{\#\{\hat\theta_n^{(b)} < \hat{\theta}_n\} / B\}$ for
$b \in \{1, \ldots, B\}$. Let $Z_{(i)}$ be the original sample without the
$i$th block $z_i$ for $i \in \{1, \ldots, k\}$, let $\hat{\theta}_{(i)}$ be
the statistic of $Z_{(i)}$, and let
$\hat{\theta}_{(.)} = \sum_{i=1}^{k} \hat{\theta}_{(i)} / k$. 
Let 
\[
\hat{a} = \frac{\sum_{i=1}^{k} (\hat{\theta}_{(.)} -
  \hat{\theta}_{(i)})^3}{6\{\sum_{i=1}^{k} (\hat{\theta}_{(.)} -
  \hat{\theta}_{(i)})^2\}^{3/2}}.
\]
Define
\[
\alpha_1 = \Phi\left(z_0 + \frac{z_{0} +
  z_{\alpha/2}}{1 - \hat{a}(\hat{z}_{0} + z_{\alpha/2})}\right)
\text{ and }
\alpha_2 = \Phi\left(z_0 + \frac{z_{0} +
  z_{1 - \alpha/2}}{1 - \hat{a}(\hat{z}_{0} + z_{1 - \alpha/2})}\right).
\]
A $(1 - \alpha)100\%~BC_a$ CI is
\[
(\hat\theta_{n, \alpha_1}^{(b)}, \quad \hat\theta_{n, \alpha_2}^{(b)}).
\]
This CI is not necessarily centered by the point estimator $\hat\theta_n$. The
$BC_a$ method corrects for bias and skewness of the $B$ bootstrap point
estimates $\{\hat\theta_n^{(1)}, \ldots, \hat\theta_n^{(B)}\}$ by including
bias-correction and acceleration factors. The acceleration factor refers to
the rate of change of the standard error of $\hat\theta_n$ with respect to
$\theta$.

@@

Three target parameters of $X_t$ were considered:
1) $\mu = 0$, the mean of $X_t$;
2) $\sigma_x = 1$, the standard deviation of $X_t$; and
2) $\phi$, the lag-1 auto-correlation coefficient.
To investigate the effect of sample size~$n$, we considered an array of values
$n \in \{100, 200, 400, 800, 1600, 3200\}$. In each configuration, we
generated 10000 replicates. For each replicate, we constructed seven 95\% block
bootstrap CIs for each parameter as described in the last section. We can
estimate $\mu$, $\sigma_x$, and $\phi$ by computing the sample mean, sample 
standard deviation, and lag-1 auto-correlation, respectively, of each
bootstrap sample, and then construct intervals for each parameter using the
appropriate procedures described in Section~\ref{sec:bbci}. Then we estimated
their actual coverage rates along with their 95\% confidence intervals from
the 10000 replicates. The block bootstrap sampling step was done with function
\texttt{tsboot} from R package \textsl{boot} \citep{boot}, with block size
$\lceil n / l \rceil$. This function by default is an implementation of moving
block bootstrap as described in the previous section, meaning that that blocks
are allowed to wrap around, and $l = \lceil n^{1/3} \rceil$.

The coverage rates of the CIs were used to compare the performance of CIs. Let
$\hat\theta_{L, r}$ and $\hat\theta_{U, r}$ be the lower and upper bound,
respectively, for the confidence interval constructed for each replicate
$r \in \{1, \ldots, R\}$, where $R$ is the number of replicates. Then
$cov = \#\{\hat\theta_{L, r} < \theta < \hat\theta_{U, r} \}/R$, for
$r \in \{1, \ldots, R\}$. If a CI method is valid, then the coverage rate is
expected to match the nominal level. Because it is unlikely for the coverage
to exactly match the nominal level, we can construct a 95\% CI of the coverage
rate (which is an estimate of a proportion with $R = 10000$. If the proportion
0.95 is included in the interval, the block bootstrap method is likely
performing well. If all values in the interval are below 0.95, the results
would suggest that the method either is providing inaccurate estimation, is
underestimating the process' variability, or a combination of both. If all
values in the interval are above .95, the results suggest that the method is
overestimating the process' variability. Figures~\ref{fig:mu}--\ref{fig:phi}
summarizes the empirical coverage rates and the 95\% confidence intervals of
the real coverage, generated using the R \textsl{ggplot2} package
\citep{ggplot2}.

@@

For estimating the mean parameter~$\mu$, Figure~\ref{fig:mu} suggests that all
methods eventually approach correct coverage of $\mu$ as sample size
increases. Student's $t$ CIs appear to need the smallest sample size
to achieve correct coverage, except for samples with strong negative
dependence, in which case, they actually over-cover $\mu$ for smaller sample
sizes. For instance, for a sample with $n = 100$ and $\phi = -0.4$, the lower
bound for a Students $t$ CIs coverage of $\mu$ is greater than 95\%, whereas
the coverage intervals for other methods contain 95\%. The standard,
Percentile, $BC$, and $BC_a$, and Proposed Method CIs require similar sample
sizes to to recover $\mu$ at the nominal level for all combinations of $n$ and
$\phi$. All methods seem to require a smaller sample to recover $\mu$ at the
nominal rate when dealing with negative dependence versus positive dependence.
For example, $BC$ CIs recovers $\mu$ for $n \geq 100$ when $\phi = -0.2$, but
they only recover $\mu$ for $n \geq 800$ when $\phi = 0.2$. In addition, as a
negative dependence gets stronger, holding everything else equal, coverage
increases, which lead to the Student $t$ CI's aforementioned over-coverage. As
a positive dependence gets stronger, holding everything else equal, coverage
decreases, and a larger sample is necessary to recover~$\mu$.

@@

This study could be used as a guide for applied statistics courses for students
to generally understand how large of a sample size is sufficient for block
bootstrap to be used versus other inference methods. For undergraduate or
graduate students, block bootstrap is not typically a part of curriculum, but
the results of this study can easily be used to demonstrate when it is
practical to use this method. This information could also prove to be useful
for research using block bootstrap estimation of time series in domains such
as econometrics. Future studies could investigate if there are types of block
bootstrap interval construction such as $ABC$ or bootstrap-t intervals, or
alternatives to block bootstrap such as AR-Sieve bootstrap that would more
\jy{Cite references for AR-sieve}
appropriately recover the parameters of a time series.
\eds{are there other methods that you know of?} \mc{mentioned an alternative
  to block bootstrap}
One could also investigate the $n$ needed to make inferences about other forms
of serially dependent data such as a moving average process.

>>>>

The goal of this paper is to provide recommendations on necessary sample size
for block bootstrap with dependent data, similar to what was done for basic
bootstrap in \citet{hesterberg2015teachers}. We consider a simple situation of
a stationary time series, where the parameters of interests are the mean,
standard deviation, and the first-order autocorrelation coefficient. We
compare six variants of block bootstrap CIs from the literature
\citep{diciccio1996bootstrap, rice2006mathematical}: a standard normal CI, and
a Student's $t$ CI, a percentile CI, a bias-corrected CI, a bias-corrected and
accelerated ($BC_a$) CI, and a CI proposed in this article. Their empirical
coverage rates at different sample sizes and dependence levels are compared in
a simulation study. The results of this study suggest that recovery of
temporal dependence parameters is reliant on the type of interval used.

@@

\paragraph{Student's $t$ CI}
The procedure for constructing a Student's $t$ CI based on standard bootstrap
is described in \citet[p.158]{efron1993introduction}. Let $t_{(\alpha, k)}$ be
the upper $\alpha$ quantile of a $t$ distribution with $k$ degrees of freedom.
With block bootstrapping, a $(1 - \alpha)100\%$ Student's $t$ CI is
\[
(\hat{\theta}_{n} - t_{(1-\alpha/2), k - 1}\hat{\text{SE}}, \quad
\hat{\theta}_{n} - t_{(\alpha/2), k -1}\hat{\text{SE}}),
\]
%\eds{define $t_{(.)}$ and check the signs}  \mc{addressed}
where $k$ is the number of blocks. This CI is centered by the point estimator
$\hat\theta_n$ and is symmetric. Like the standard normal interval, the
Student's $t$ CI is a table-based bootstrap interval approach. In this case,
its validity relies on whether the distribution of $\hat\theta_n$ is
reasonably well approximation by the $t_{k-1}$ distribution with an expected
value of $\theta$, and whether the bootstrap $\widehat{\text{SE}}$
approximates the true standard error.

@@

\paragraph{Bias Corrected and Accelerated ($BC_a$) Bootstrap CI}
The $BC_a$ CI was first suggested in \citet{efron1987better}. Let
$\hat{z}_0 = \Phi^{-1}\{\#\{\hat\theta_n^{(b)} < \hat{\theta}_n\} / B\}$ for
$b \in \{1, \ldots, B\}$. Let $Z_{(i)}$ be the original sample without the
$i$th block $z_i$ for $i \in \{1, \ldots, k\}$, let $\hat{\theta}_{(i)}$ be
the statistic of $Z_{(i)}$, and let
$\hat{\theta}_{(.)} = \sum_{i=1}^{k} \hat{\theta}_{(i)} / k$. 
Let 
\[
\hat{a} = \frac{\sum_{i=1}^{k} (\hat{\theta}_{(.)} -
  \hat{\theta}_{(i)})^3}{6\{\sum_{i=1}^{k} (\hat{\theta}_{(.)} -
  \hat{\theta}_{(i)})^2\}^{3/2}}.
\]
Define
\[
\alpha_1 = \Phi\left(z_0 + \frac{z_{0} +
  z_{\alpha/2}}{1 - \hat{a}(\hat{z}_{0} + z_{\alpha/2})}\right)
\text{ and }
\alpha_2 = \Phi\left(z_0 + \frac{z_{0} +
  z_{1 - \alpha/2}}{1 - \hat{a}(\hat{z}_{0} + z_{1 - \alpha/2})}\right).
\]
A $(1 - \alpha)100\%~BC_a$ CI is
\[
(\hat\theta_{n, \alpha_1}^{(b)}, \quad \hat\theta_{n, \alpha_2}^{(b)}).
\]
This CI is not necessarily centered by the point estimator $\hat\theta_n$. The
$BC_a$ method corrects for bias and skewness of the $B$ bootstrap point
estimates $\{\hat\theta_n^{(1)}, \ldots, \hat\theta_n^{(B)}\}$ by including
bias-correction and acceleration factors. The acceleration factor refers to
the rate of change of the standard error of $\hat\theta_n$ with respect to
$\theta$.

@@

Three target parameters of $X_t$ were considered:
1) $\mu = 0$, the mean of $X_t$;
2) $\sigma_x = 1$, the standard deviation of $X_t$; and
3) $\phi$, the lag-1 auto-correlation coefficient.
To investigate the effect of sample size~$n$, we considered an array of values
$n \in \{100, 200, 400, 800, 1600, 3200\}$. In each configuration, we
generated 10000 replicates. For each replicate, we constructed seven 95\% block
bootstrap CIs for each parameter as described in the last section. We can
estimate $\mu$, $\sigma_x$, and $\phi$ by computing the sample mean, sample 
standard deviation, and lag-1 auto-correlation, respectively, of each
bootstrap sample, and then construct intervals for each parameter using the
appropriate procedures described in Section~\ref{sec:bbci}. Then we estimated
their actual coverage rates along with their 95\% confidence intervals from
the 10000 replicates. The block bootstrap sampling step was done with function
\texttt{tsboot} from R package \textsl{boot} \citep{boot}, with block size
$\lceil n / l \rceil$. This function by default is an implementation of moving
block bootstrap as described in the previous section, meaning that that blocks
are allowed to wrap around, and $l = \lceil n^{1/3} \rceil$.

The coverage rates of the CIs were used to compare the performance of CIs. Let
$\hat\theta_{L, r}$ and $\hat\theta_{U, r}$ be the lower and upper bound,
respectively, for the confidence interval constructed for each replicate
$r \in \{1, \ldots, R\}$, where $R$ is the number of replicates. Then
$cov = \#\{\hat\theta_{L, r} < \theta < \hat\theta_{U, r} \}/R$, for
$r \in \{1, \ldots, R\}$. If a CI method is valid, then the coverage rate is
expected to match the nominal level. Because it is unlikely for the coverage
to exactly match the nominal level, we can construct a 95\% CI of the coverage
rate (which is an estimate of a proportion with $R = 10000$). If the proportion
0.95 is included in the interval, the block bootstrap method is likely
performing well. If all values in the interval are below 0.95, the results
would suggest that the method either is providing inaccurate estimation, is
underestimating the process' variability, or a combination of both. If all
values in the interval are above .95, the results suggest that the method is
overestimating the process' variability. Figures~\ref{fig:mu}--\ref{fig:phi}
summarizes the empirical coverage rates and the 95\% confidence intervals of
the real coverage, generated using the R \textsl{ggplot2} package
\citep{ggplot2}.

@@

For estimating the mean parameter~$\mu$, Figure~\ref{fig:mu} suggests that all
methods eventually approach correct coverage of $\mu$ as sample size
increases. Student's $t$ CIs appear to need the smallest sample size
to achieve correct coverage, except for samples with strong negative
dependence, in which case, they actually over-cover $\mu$ for smaller sample
sizes. For instance, for a sample with $n = 100$ and $\phi = -0.4$, the lower
bound for a Students $t$ CIs coverage of $\mu$ is greater than 95\%, whereas
the coverage intervals for other methods contain 95\%. The standard,
Percentile, $BC$, and $BC_a$, and Proposed Method CIs require similar sample
sizes to recover $\mu$ at the nominal level for all combinations of $n$ and
$\phi$. All methods seem to require a smaller sample to recover $\mu$ at the
nominal rate when dealing with negative dependence versus positive dependence.
For example, $BC$ CIs recover $\mu$ for $n \geq 100$ when $\phi = -0.2$, but
they only recover $\mu$ for $n \geq 800$ when $\phi = 0.2$. In addition, as a
negative dependence gets stronger, holding everything else equal, coverage
increases, which lead to the Student $t$ CI's aforementioned over-coverage. As
a positive dependence gets stronger, holding everything else equal, coverage
decreases, and a larger sample is necessary to recover~$\mu$.

@@

This study could be used as a guide for applied statistics courses for students
to generally understand how large of a sample size is sufficient for block
bootstrap to be used versus other inference methods. For undergraduate or
graduate students, block bootstrap is not typically a part of curriculum, but
the results of this study can easily be used to demonstrate when it is
practical to use this method. This information could also prove to be useful
for research using block bootstrap estimation of time series in domains such
as econometrics. Future studies could investigate if there are types of block
bootstrap interval construction such as $ABC$ or bootstrap-t intervals, or
alternatives to block bootstrap such as AR-Sieve bootstrap that would more
\jy{Cite references for AR-sieve}
appropriately recover the parameters of a time series.

\eds{Check out this paper:
Peter Buhlmann. "Bootstraps for Time Series." Statist. Sci. 17(1) 52 - 72, 
May 2002 - the conclusions of this paper are interesting in their own right:
``Disadvantages of the [block bootstrap] method include the following.
The block bootstrap sample should not be viewed as
a reasonable sample mimicking the data-generating
process: it is not stationary and exhibits artifacts where
resampled blocks are linked together. This implies that
the plug-in rule for bootstrapping an estimator $\hat{\theta}$ is not
appropriate. A prevectorization of the data is highly
recommended, but the bootstrapped estimator and its
computing routine may then need to be redesigned. As
a general nonparametric scheme, the block bootstrap
may be outperformed in various niches of stationary
time series, for example, for linear time series (see Section 3) 
and for categorical processes (see Section 5).
Second-order accuracy for a confidence interval has
been justified with the approach of Studentizing and
BCa correction (in the case of noncategorical time series); 
the latter was found to yield marginal improvement in a 
simulated example (we did not consider the former). }

One could also investigate the $n$ needed to make inferences about other forms
of serially dependent data such as a moving average process.
